---
title: "IPW illustration report"
author: "Lucas B. de Azevedo"
date: "2026-06-15"
output: html_document
---

# Packages
```{r installing packages}
#The installing step is not necessary if you use the renv file, available in
# the project's repository and copy the complete environment in which the code
# was executed

install.packages("nhanesA")
install.packages("tidyverse")
install.packages("gtsummary")
install.packages("WeighIt")
install.packages("cobalt")
install.packages("sandwich")
install.packages("lmtest")
```

# Using nhanesA package to access the database and do the variable selection

## Library
```{r library for nhanes data}
library(nhanesA) # You may access package documentation through ??nhanesA
library(tidyverse) # or ?tidyverse. It does apply for futher packages info
```

## Obtaining information about selected variables
```{r nhanes variables}
nhanesCodebook("DEMO_J",     #Demographic information
               colname = c(
                 "SEQN",     #Identification (ID)
                 "RIAGENDR", #Sex
                 "RIDAGEYR", #Age
                 "RIDRETH3" #Ethnicity 
                 )
               )

nhanesCodebook("MCQ_J",    #Medical related information
               colname = c(
                 "SEQN",   #ID
                 "MCQ010", #Ever been told you have asthma
                 "MCQ025", #Age when first had asthma
                 "MCQ035", #Still have asthma
                 "MCQ040"  #Had asthma attack in past year
                 )
               )

nhanesCodebook("SMQ_J",     #Smoking related questions
               colname = c(
                 "SEQN",    #ID
                 "SMQ020",  #Smoked at least 100 cigarettes in life
                 "SMQ040"   #Do you now smoke cigarettes?
                 )
               )

nhanesCodebook("BMX_J",   #Body measures
               colname = c(
                 "SEQN",  #ID
                 "BMXBMI" #Body Mass Index (BMI)
                 )
               )
```

## Creating objects and merging information
```{r nhanes data frame}
demo <- nhanes("DEMO_J") |> 
  select(SEQN, RIAGENDR, RIDAGEYR, RIDRETH3) #Selecting variables

mcq <- nhanes("MCQ_J") |>
  select(SEQN, MCQ010, MCQ025, MCQ035, MCQ040)

smq <- nhanes("SMQ_J") |>
  select(SEQN, SMQ020, SMQ040)

bmx <- nhanes("BMX_J") |>
  select(SEQN, BMXBMI)

# Merging data frame

df_raw <- demo |> 
  left_join(mcq, by = "SEQN") |> #Joins variables using a reference
  left_join(smq, by = "SEQN") |> 
  left_join(bmx, by = "SEQN")
```

## Modifying varibles to perform analysis
```{r nhanes df variables modification}
df_asm <- df_raw |>
  filter(MCQ035 == "Yes") |>                          #Active asthma 
  filter(SMQ040 %in% c("Every day", 
                        "Some days", 
                        "Not at all")) |>             
  mutate(
    A             = case_when(
      SMQ040    %in% c("Every day", "Some days")   ~ 1L,  #Current smoker
      SMQ040      == "Not at all"                  ~ 0L,  
      TRUE                                         ~ NA_integer_
    ),
    Y             = case_when(
      MCQ040      == "Yes" ~ 1L, #"Asthma attack"
      MCQ040      == "No"  ~ 0L,
      TRUE                 ~ NA_integer_
    ),
    sex           = factor(RIAGENDR),                 #Transfoming in factors
    race          = factor(RIDRETH3),
    age           = RIDAGEYR,
    bmi           = BMXBMI,
    infant_asm    = ifelse(MCQ025 <= 12, 1L, 0L) #If asthma with less than 12yr
  ) |>                                           # = 1, 0 otherwise
  drop_na(A, Y, sex, race, age, bmi)
```

## Summarizing
```{r gtsummary to table creation}
library(gtsummary)

table_asm <- 
  df_asm |> 
  tbl_summary(
    include = c(
      sex, age, race, bmi, A
    ),
              by = Y,
              label = list(
                sex        = "Sex",
                age        = "Age (years)",
                race       = "Race",
                bmi        = "Body Mass Index (kg/m²)",
                infant_asm = "Asthma before 13 years old",
                A          = "Current smoking"
              ),
              statistic = list(
                all_continuous()  ~ "{mean} ({sd})", 
                all_categorical() ~ "{n} ({p}%)"
                ) 
              ) |> 
  add_difference(
    test = everything() ~ "smd" # Difference by standard mean
    ) |> 
  modify_header(
    label    = "**Variable**",
    stat_1   = "**No asthma attack**  \nN = 124",
    stat_2   = "**Asthma attack**  \nN = 108",
    estimate = "**SMD**"
    ) |>
  modify_column_hide(
    columns = c(
      conf.low, conf.high
      )
    ) 
  #Hiding IC column that automatically display confidence intervals when
  #using add_difference()

table_asm
```

# Inverse Probability Weight (IPW)
```{r creating weights using WeightIt}
library(WeightIt)

w_out <- weightit(
  A ~ sex + age + race + bmi + infant_asm, #Regression formula (PS)
  data    = df_asm,
  method  = "glm",
  estimand = "ATE",
  stabilize = TRUE #SW(A) - Stabilized Weights
)

summary(w_out) #Cheking if there is any explosive weights

mean(w_out$weights) #Checking model specification
```

# Assessing balance applying weights
```{r assessing balance after IPW}
library(cobalt)
library(tidyverse)

bal.tab(w_out,
        thresholds = c(m = 0.1), #SMD threshold of 0.1
        un         = TRUE)

#Setting up variable names
v <- data.frame(old = c("age", "sex_Female", "race_Mexican American", 
                        "race_Other Hispanic", "race_Non-Hispanic White", 
                        "race_Non-Hispanic Black", "race_Non-Hispanic Asian", 
                        "race_Other Race - Including Multi-Racial", "bmi", 
                        "infant_asm"
                        ),
                new = c("Age", "Female", "Mexican American", "Other Hispanic", 
                        "Non-Hispanic White", "Non-Hispanic Black", 
                        "Non-Hispanic Asian", "Other", "BMI", 
                        "Asthma before 13 years old"
                        )
                )

love.plot(w_out,
          thresholds  = c(m = 0.1),
          colors      = c("#FF0000", "#21908C"),  # viridis colors codes
          shapes      = c(16, 17),
          var.order   = "unadjusted",
          var.names   = v
          ) 
```

# Adjusting the Marginal Structured Model (MSM) after confounding balance
```{r MSM using robust variace}
library(sandwich)
library(lmtest)

msm_asma <- glm(
  Y ~ A,
  data    = df_asm,
  family  = binomial(link = "logit"),
  weights = w_out$weights   
)

coeftest(msm_asma,
         vcov = vcovHC(
           msm_asma, type = "HC3" #Robust variance
           )
         )

# Exponenciating to transform coeficients in OR
exp(coef(msm_asma)["A"])

exp(
  confint.default(
  msm_asma,
  parm = "A",
  vcov = vcovHC(
    msm_asma,
    type = "HC3"
    )
  )
)
```